---
title: "PA1_template"
author: "Brigitte Vogelsangs"
date: "24-1-2020"
output: 
  html_document: 
    keep_md: yes
---

```{r setup}
library(knitr)
knitr::opts_chunk$set(echo=TRUE)
```

# Course project 1

## Loading and preprocess the data
The code used to load the data. 
Dates are processed and summery created.

```{r activity}
library(dplyr)
library(ggplot2)

activity <- read.csv("activity.csv")
activity$date <- as.POSIXct(activity$date, tz= "GMT", "%Y-%m-%d")
weekday <- weekdays(activity$date)
activity <- cbind(activity, weekday)
summary(activity)
```

## What is the mean total number of steps taken per day?

Missing values in the dataset are ignored.

- Calculate the total number of steps taken per day
- Make a histogram of the total number of steps taken each day
- Calculate and report the mean and median of the total number of steps taken per day

```{r activity_steps_day, echo=TRUE}
activity_steps_day <- with(activity, aggregate(steps, by = list(date), FUN = sum, na.rm = TRUE))
names(activity_steps_day) <- c("date", "steps")
hist1 <- hist(activity_steps_day$steps, main = "Total number of steps taken per day", 
     xlab = "Total steps taken per day", col = "blue", ylim = c(0,20), 
     breaks = seq(0,25000, by=2500))
mean(activity_steps_day$steps)
median(activity_steps_day$steps)
```

## What is the daily average activity pattern?

- Make a time series plot (i.e. \color{red}{\verb|type = "l"|}type="l") of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all days (y-axis)
- Which 5-minute interval, on average across all the days in the dataset, contains the maximum number of steps?

```{r activity_day_average}
activity_day_average <- aggregate(activity$steps, by=list(activity$interval), FUN=mean, na.rm=TRUE)
names(activity_day_average) <- c("interval", "mean")
plot10 <- plot(activity_day_average$interval, activity_day_average$mean, type = "l", 
               col="blue", lwd = 2, xlab="Interval", ylab="Average number of steps", 
               main="Average number of steps per intervals")
activity_day_average[which.max(activity_day_average$mean), ]$interval
```

## Imputing missing values
There are a number of days/intervals where there are missing values (coded as \color{red}{\verb|NA|}NA). 
The presence of missing days may introduce bias into some calculations or summaries of the data.

Calculate and report the total number of missing values in the dataset.

Created a set without missing data activity new by using the mean for the day


```{r activity_replaceNA}
sum(is.na(activity$steps))
activity_replaceNA <- activity_day_average$mean[match(activity$interval, activity_day_average$interval)]
```

## Create a new dataset that is equal to the original dataset but with the missing data filled in.
- Make a histogram of the total number of steps taken each day
- Calculate and report the mean and median total number of steps taken per day. 
- Do these values differ from the estimates from the first part of the assignment? 
- What is the impact of imputing missing data on the estimates of the total daily number of steps?

```{r activity_new}
activity_new <- transform(activity, steps = ifelse(is.na(activity$steps), yes = activity_replaceNA, no = activity$steps))
total_steps_new <- aggregate(steps ~ date, activity_new, sum)
names(total_steps_new) <- c("date", "daily_steps")
hist2 <- hist(total_steps_new$daily_steps, col = "blue", 
              xlab = "Total steps per day", ylim = c(0,30), 
              main = "Total number of steps taken each day", breaks = seq(0,25000,by=2500))
mean(total_steps_new$daily_steps)
median(total_steps_new$daily_steps)
```

There is a difference in value between the mean and the median
```{r activity_difference}
difference1 <- mean(activity_steps_day$steps)-mean(total_steps_new$daily_steps)
difference2 <- median(activity_steps_day$steps)-median(total_steps_new$daily_steps)
difference1
difference2
```
The impact shows a higher frequency of the total number of steps taken each day.

## Are there differences in activity patterns between weekdays and weekends?

- Create a new factor variable in the dataset with two levels – “weekday” and “weekend” indicating whether a given date is a weekday or weekend day.

```{r activity_weekdays_weekends}
activity_new$date <- as.Date(strptime(activity_new$date, format="%Y-%m-%d"))
activity_new$datetype <- sapply(activity_new$date, function(x) {
        if (weekdays(x) == "zaterdag" | weekdays(x) =="zondag") 
                {y <- "Weekend"} else 
                {y <- "Weekday"}
                y
        })
str(activity_new$date)
str(activity_new$datetype)
```

- Make a panel plot containing a time series plot of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all weekday days or weekend days (y-axis). 

```{r activity_time_series}
activity_new$date <- as.Date(strptime(activity_new$date, format="%Y-%m-%d"))
activity_new$datetype <- sapply(activity_new$date, function(x) {
        if (weekdays(x) == "zaterdag" | weekdays(x) =="zondag") 
                {y <- "Weekend"} else 
                {y <- "Weekday"}
                y
        })
activity_time_series <- aggregate(steps~interval + datetype, activity_new, mean, na.rm = TRUE)
plot11 <- ggplot(activity_time_series, aes(x = interval , y = steps, color = datetype)) +
    geom_line() + labs(title = "Average daily steps by type of date", x = "Interval", 
                       y = "Average number of steps") + facet_wrap(~datetype, ncol = 1, nrow=2)
print(plot11)
```

